Type-Safe Modular Parsing

### 2017-01-01 paper/sections/Inheritance.tex
Revised S4.

Unfortunately, it fails to parse some expressions like \lstinline{"1 + x"}, which is obviously valid in the new grammar.
The reason is that \lstinline{pAdd} makes two recursive calls to parse sub-expressions, by using \lstinline{pExpr}, which
covers both cases in the old grammar. But the newly added case \lstinline{pVar} is not observed by the recursive \lstinline{pExpr},
hence the parser does not work as expected. Still, one may continue to change the recursive call in \lstinline{pAdd} with \lstinline{pExtExpr},
but modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

@@

\lstinputlisting[linerange=68-73]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=INHERITANCE_APPROACH
Now \lstinline{ExtParser} successfully represents the parser for the extended language, because Scala uses dynamic dispatch for
method overriding in inheritance. When the input \lstinline{"1 + x"} is fed to the parser \lstinline{this.pExpr}, it firstly delegates
the work to \lstinline{super.pExpr}, which parses literals and additions. However, the recursive call \lstinline{pExpr} in \lstinline{pAdd}
actually refers to \lstinline{this.pExpr} again, due to dynamic dispatch. It implies that the recursive call has included the extension. This
elegant inheritance-based approach opens the recursion of parsers in a modular way.

@@

\subsection{Object Algebras for full Extensibility}\label{subsec:overview-oa}

And also parser for it as below. Notice the parsing function \inlinecode{pExpr} now takes an argument of type \inlinecode{ExprAlg[E]}, which means it accepts any instance of \inlinecode{ExprAlg}. Such an instance can for example be \lstinline{Print}, which takes charge of operations on expressions, and parsing code only does delegation.

\lstinputlisting[linerange=29-34]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PARSER

\haoyuan{I'm trying to avoid using terminology like algebra and algebra interface.}

\paragraph{Extensibility of Syntax} Following previous examples, the code below shows how to extend the language with variables.

\lstinputlisting[linerange=38-51]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT
Note that we use the new combinator \lstinline{<|>} defined in our library, to compose two parsing functions into one. The function \lstinline{pExtExpr} has the compound type \lstinline{ExprAlg[E] with VarAlg[E]} in its argument, where two language ASTs are combined to represent the new language.

To use the parser, we must provide an algebra instance as the operation to construct the parsing results. In the code below, we use the combination of \lstinline{Print} and \lstinline{PrintVar}, so that the parsing result is a pretty printing of the AST.

\lstinputlisting[linerange=60-65]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

\paragraph{Extensibility of Operations} With Object Algebras, operations over ASTs can also be extended in a modular way. Here is an example of collecting free variables from an expression. We can feed this operation to the parser and obtain a set of free variables.

\lstinputlisting[linerange=79-85]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

\haoyuan{I think in our library language ASTs are composed using intersection, not inheritance.}

\paragraph{Multiple Sorts of Syntax} Another advantage of using Object Algebras is that it supports multiple sorts of syntax easily. In several cases, we want to divide the syntactic elements into some groups. For example, the grammar below has two sorts, which are expressions and types.

\begin{tabular}{m{0.45\linewidth}m{0.45\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::=  `\\' <ident> `:' <type> `.' <expr>
\end{grammar}
\end{tabular}

Using Object Algebras, we can easily distinguish them just by adding an extra type parameter. The abstract syntax is
presented below:

\lstinputlisting[linerange=70-74]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
In the next sections, we will discuss them with more details.

>>>>

Unfortunately, it fails to parse some expressions like \lstinline{"1 + x"}, which is obviously valid in the new grammar.
The reason is that \lstinline{pAdd} makes two recursive calls to parse sub-expressions, by using \lstinline{pExpr}, which
covers both cases in the old grammar. But the newly added case \lstinline{pVar} is not observed by the recursive \lstinline{pExpr},
hence the parser does not work as expected. Still, one may continue to change the recursive call in \lstinline{pAdd} with \lstinline{pVarExpr},
but modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

@@

\lstinputlisting[linerange=68-73]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=INHERITANCE_APPROACH
Now \lstinline{VarExprParser} successfully represents the parser for the extended language, because Scala uses dynamic dispatch for
method overriding in inheritance. When the input \lstinline{"1 + x"} is fed to the parser \lstinline{this.pExpr}, it firstly delegates
the work to \lstinline{super.pExpr}, which parses literals and additions. However, the recursive call \lstinline{pExpr} in \lstinline{pAdd}
actually refers to \lstinline{this.pExpr} again, due to dynamic dispatch. It implies that the recursive call has included the extension. This
elegant inheritance-based approach opens the recursion of parsers in a modular way.

@@

### 2017-01-01 paper/sections/ObjectAlgebras.tex
Revised S4.

\lstinputlisting[linerange=8-11]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_ALG

The language has two constructors: literals and additions. The trait \lstinline{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

\lstinputlisting[linerange=15-18]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PRINT

Here \lstinline{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
Furthermore, one can define a new trait to implement evaluation on arithmetic expressions:

\lstinputlisting[linerange=18-21]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_EXPR_PRINT

On the other hand, the data variants can be extended by inheriting \lstinline{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \lstinline{VarAlg} is defined as follows:\haoyuan{Updates till here.}

\lstinputlisting[linerange=25-27]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT

Now pretty-printing on the new language can be realized by code reuse, without modifying existing code:

\lstinputlisting[linerange=31-33]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_LIT_PRINT

On this, to create an expression of \lstinline{LitAlg}, a generic method is defined as follows:

\huang{todo: better name for build}
\lstinputlisting[linerange=47-47]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_BUILD

Such a method implicitly represents the expression \lstinline{"x 3"}. The code

\begin{lstlisting}
build(new LitPrint{})
\end{lstlisting}

results in \lstinline{"(x 3)"}, as the result of pretty-printing. Now we observe that an ``object'' of expression actually
has the function type \lstinline{LitAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform an \lstinline{ExprAlg} expression to a \lstinline{LitAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:

\lstinputlisting[linerange=37-41]{../Scala/Parser/src/PaperCode/ObjectAlg/OA1.scala}% APPLY:linerange=OBJECT_ALG_REFACTOR

When \lstinline{env = Map("y"} \lstinline{-> 3)}, the \lstinline{ExprAlg} expression \lstinline{"x y"} is changed to \lstinline{LitAlg} expression \lstinline{"x 3"}. The abstract method
\lstinline{alg} in \lstinline{Refactor} can be any algebra of \lstinline{SubAlg}, simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to apply transformation to an \lstinline{ExprAlg} expression before \lstinline{alg} returns the result.

@@

Just as shown above, \lstinline{ExprAlg} is defined for a small language in an Object-Algebra style. Furthermore,
A parser can be defined for this whole language at once. Since Object Algebras represent ``objects'' implicitly as functions like \lstinline{ExprAlg[E] =>} \lstinline{E} for generic \lstinline{E}, such a parser should consume an algebra from its parameter, then return a value of \lstinline{Fix[Parser[E]]} as in open recursion, where \lstinline{E} is abstract.
\begin{lstlisting}
trait ExprParser[E] {
    val pE : ExprAlg[E] => Fix[Parser[E]] = alg => p =>
        ident ^^ alg.varE |||
        p ~ p ^^ { case e1 ~ e2 => alg.appE(e1, e2) }
}
\end{lstlisting}
It is observed that \lstinline{pE} is in fact a parser generator, where \lstinline{alg} appears as a parameter, which can be any algebra, and after \lstinline{alg} is fed, it returns a \lstinline{Fix[Parser[E]]}. Again \lstinline{p} is the explicit self-reference of the open parser. Inside the body, the algebra is invoked correspondingly for all cases right after parsing. Note that the two cases are combined using the original \lstinline{|||} operator. To make \lstinline{alg} as a generic algebra, the enclosing trait of the parser, namely \lstinline{ExprParser}, is abstracted over \lstinline{E}. Behaviors, or algebras that can be fed to such a parser, are expected to be defined independently from the Object Algebra interface, just like the algebra \lstinline{ExprPrint} in Section~\ref{subsec:objectalgebras}.

To explain the modularity, we again add literals to expressions, but instead of using inheritance as Section~\ref{subsec:objectalgebras}, \lstinline{LitAlg} is defined independently together with its own parser and algebra:
\begin{lstlisting}
trait LitAlg[E] {
    def litE(n : Int) : E
}

trait LitPrint {
    def litE(n : Int) = n.toString
}

trait LitParser[E] {
    val pE : LitAlg[E] => Fix[Parser[E]] = alg => p =>
        numericLit ^^ alg.litE
}
\end{lstlisting}
In this case, \lstinline{LitAlg} is yet another small language, which is self-contained. To merge the two languages together, we require their grammars
to be merged, and we obtain a combined language, whose parser integrates the two small parsers using alternative. In Scala we use compound types, namely use \lstinline{"with"} for two Object Algebra interface types, to avoid polluting the namespace.

\begin{lstlisting}
trait ExprLitParser[E] {
    val pE : (ExprAlg[E] with LitAlg[E]) => Fix[Parser[E]] = alg => p => {
        val pExprE = new ExprParser[E]{}.pE(alg)(p)
        val pLitE = new LitParser[E]{}.pE(alg)(p)
        pExprE ||| pLitE
    }
}
\end{lstlisting}

In client code, a user can either define a new function with type \lstinline{(ExprAlg[E] with LitAlg[E])} \lstinline{=>} \lstinline{Parser[E]} based on \lstinline{pE}, which implies an algebra is applied after \lstinline{fix}, or simply feed an algebra to \lstinline{pE} at first, then obtain the result from \lstinline{fix}:
\begin{lstlisting}
val parsePrint = fix(new ExprLitParser[String]{}.pE(new ExprPrint with LitPrint{}))
val result = runParser(parsePrint)("x 3") // prints "(x 3)"
\end{lstlisting}

@@

One intuition of using Object Algebras, as illustrated above, is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, or realize different operations on structures, without modifying existing code, meanwhile separate compilation is supported. Another one is that Object Algebras benefit us from using generics, where we can abstract multiple syntax in the type parameters, and define less classes, simplifying code comparatively. Below we refactor the last example in Section~\ref{subsec:parsingwithopen}, by defining a new Object Algebra interface with two cases \lstinline{intT} and \lstinline{lamE}. Meanwhile a parser is implemented for that language.
\begin{lstlisting}
trait LamAlg[E, T] {
    def intT() : T
    def lamE(x : String, t : T, e : E) : E
}

trait LamParser[E, T] {
    type Pair = (Parser[E], Parser[T])
    val pE : LamAlg[E, T] => (=> Pair) => Parser[E] = alg => p =>
        ("\\" ~> ident) ~ (":" ~> p._2) ~ ("." ~> p._1)
            ^^ { case x ~ t ~ e => alg.lamE(x, t, e) }
    val pT : LamAlg[E, T] => (=> Pair) => Parser[T] = alg => p =>
        "Int" ^^^ alg.intT
}
\end{lstlisting}
Note that we are omitting its lexer here. \lstinline{LamParser} contains parsers for both expressions and types. From the code we observe that the multi-sorts of Object Algebra interface
exactly express multiple syntax. At this point, to combine these two small components together we need some ``glue code''. But instead of writing composition by hand, we reimplement the combinator \lstinline{|||}, to compose one parser with type \lstinline{Alg1 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} and another one with type \lstinline{Alg2 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} using alternative. Note that we can only abstract over the whole algebra interface type, instead of working on its type parameters, as the number of type parameters is not fixed. In summary, \lstinline{"|||"} produces a combined parser, which takes a compound algebra as its parameter:
\begin{lstlisting}
implicit class Combinator[Alg1, Alg2, R, E](x : Alg1 => (=> R) => Parser[E]) {
    def |||(y : Alg2 => (=> R) => Parser[E]): (Alg1 with Alg2) => (=> R) => Parser[E]
        = alg => l => x(alg)(l) ||| y(alg)(l)
}
\end{lstlisting}
Now it is sufficient to merge \lstinline{ExprParser} and \lstinline{LamParser} together:
\begin{lstlisting}
trait ExprLamParser[E, T] {
    val pExpr = new ExprParser[E]{}
    val pLam = new LamParser[E, T]{}
    val pExprE : ExprAlg[E] => (=> pLam.Pair) => Parser[E] =
        alg => p => pExpr.pE(alg)(p._1)

    val pE = pExprE ||| pLam.pE
    val pT = pLam.pT
}
\end{lstlisting}
Parsers for expressions and types are connected using \lstinline{|||}, respectively. Hence the isolation for different syntax
is guaranteed by the type system. The only left problem is that we still need to replace the parser in \lstinline{ExprParser} with \lstinline{pExprE}
by updating the fix-point parameter. In the next section we will address this problem, and meanwhile our implementation enhances modularity in various aspects.

>>>>

\lstinputlisting[linerange=8-11]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_ALG
The language has two constructors: literals and additions. The trait \lstinline{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

\lstinputlisting[linerange=15-18]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \lstinline{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
Furthermore, one can define a new trait to implement evaluation on arithmetic expressions:

\lstinputlisting[linerange=22-25]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EVAL

On the other hand, the data variants can be extended by inheriting \lstinline{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \lstinline{VarAlg} is defined as follows:

\lstinputlisting[linerange=29-31]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_ALGEXT
Now pretty-printing on the new language can be realized without modifying existing code:

\lstinputlisting[linerange=35-37]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
We observe that only the new case is implemented for pretty-printing, and the others have been inherited.
It offers programmers with great code reuse, in the sense that existing cases can also be
overridden in \lstinline{VarExprPrint} using \lstinline{"override def"}.

On this, to create an expression of \lstinline{VarExprAlg}, a generic method is defined as follows:

\huang{todo: better name for build}\haoyuan{makeExp}
\lstinputlisting[linerange=41-41]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP
Such a method implicitly represents the expression \lstinline{"1 + x"}. The code

\begin{lstlisting}
makeExp(new VarExprPrint {})
\end{lstlisting}
results in \lstinline{"(1 + x)"}, as the result of pretty-printing. Now we observe that an ``object'' of expression actually
has the function type \lstinline{VarExprAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform a \lstinline{VarExprAlg} expression to an \lstinline{ExprAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:\haoyuan{mention in discussion that shy deals with boilerplate.}

\lstinputlisting[linerange=45-56]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_REFACTOR
Here the \lstinline{VarExprAlg} expression \lstinline{"1 + x"} is changed to \lstinline{ExprAlg} expression \lstinline{"1 + 2"}, before the pretty-printing algebra is applied. The abstract method
\lstinline{alg} in \lstinline{Refactor} can actually be any algebra simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to transform a \lstinline{VarExprAlg} expression before we obtain the result.

@@

Just as shown above, \lstinline{ExprAlg} is defined for a small language in an Object-Algebra style. Furthermore,
A parser can be defined for this whole language at once. Since Object Algebras represent ``objects'' implicitly as functions like \lstinline{ExprAlg[E] =>} \lstinline{E} for generic \lstinline{E}, such a parser should have type \lstinline{ExprAlg[E] =>} \lstinline{Parser[E]}, where \lstinline{E} can be abstracted by the enclosing type.

\lstinputlisting[linerange=60-68]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, yet it does not appear to be extensible. For example, we have demonstrated in Section~\ref{subsec:inheritance-approach} that method overriding can update \lstinline{pExpr} for an extended syntax, but now \lstinline{pExpr} will have a different type, like \lstinline{VarExprAlg[E] =>} \lstinline{Parser[E]} where the difference appears in the argument. The difference of types prohibits method overriding, however, this problem refers to \textit{binary methods}. \haoyuan{Shall we mention this?}

One solution to this problem is taking out the argument as a single field, hence it supports overriding as well. Therefore we refactor the parsing code as follows:
\lstinputlisting[linerange=72-81]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing. Note that \lstinline{pE} is the one that is used recursively, exported by the trait and considered to be overridden from inheritance. The reason why we use an extra and seemingly redundant \lstinline{pExprE}, is that \lstinline{pE} can no longer be referred to by the \lstinline{super} keyword, as a limitation of Scala for generics, hence we can directly invoke \lstinline{pExprE} as the specific \lstinline{pE} in \lstinline{ExprOAParser}. And we still want code reuse for the composition of \lstinline{pLit} and \lstinline{pAdd}. Now it is worth trying on this pattern with an extension, like \lstinline{VarExprAlg} which we have been using:
\lstinputlisting[linerange=85-96]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=EXT_OA_PARSER
In the client code shown above, an instance of algebra is fed to \lstinline{VarExprOAParser} for pretty-printing expressions. Then the trait exports \lstinline{pE}, parses the input and applies the algebra. The result shows that such a pattern is a nice solution to modular parsing. At this point, one may argue that with a different operation like evaluation, the redundant code is required to obtain the parser, and it results in parsing the same input twice. This is generally solved in the original Object Algebras paper [] by composing two algebras into one in advance.

@@

One intuition of using Object Algebras, as illustrated above, is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, or realize different operations on structures, without modifying existing code, meanwhile separate compilation is supported. Another one is that Object Algebras benefit us from using generics, where we can abstract multiple syntax in the type parameters, and define less classes, simplifying code comparatively. Below we further extend the language with types, by introducing an extra type parameter in the algebra interface. The extended grammar is:\\

\begin{tabular}{m{0.45\linewidth}m{0.45\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= ... \alt `int' \alt <type> `->' <type>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::=  ... \alt `\\' <ident> `:' <type> `.' <expr>
\end{grammar}
\end{tabular}

Hence we write code for the extended AST, pretty-printing operation and the extended parser:

\lstinputlisting[linerange=100-128]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
Besides lexing, the trait \lstinline{TypedLamOAParser} also introduces parsers for types, and the new case for expressions. For the same reason,
we use \lstinline{pTypedLamT} and \lstinline{pTypedLamE} as copies of current \lstinline{pT} and \lstinline{pE}, recursively. \lstinline{pT} and \lstinline{pE} are actually the parsers for recursion.

From the code we observe that the multi-sorts of Object Algebra interface
exactly express multiple syntax. The isolation for different syntax
is guaranteed by the type system. Now it is straightforward to show some client code
for the new parser:

\lstinputlisting[linerange=132-134]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
Pretty-printing the parsed results just works as expected.
